---
title: "Tidymodels for Classification"
author: "Ben Stockton"
format: 
    html:
        theme: flatly
        toc: true
        toc-location: left
        code-fold: false
        highlight-style: github
        fig-align: center
        number-sections: true
    revealjs:
        output-file: tree_class_slides
        theme: simple
        slide-number: true
        chalkboard: true
        code-fold: false
        highlight-style: github
        fig-align: center
        incremental: false
# bibliography: references.bib
# csl: https://raw.githubusercontent.com/citation-style-language/styles/master/apa.csl
execute:
  echo: true
  warning: false
  message: false
  cache: true
---

# Introduction

This lab explores tree based methods - decision trees and random forest - for classification, i.e. prediction of binary outcome.

## Set-up

```{r}
#| label: set-up
library(dplyr)
library(ggplot2)
library(caret)
library(randomForest)
library(rpart)
library(rpart.plot)
library(tidymodels)
library(vip)
```


---

## 🫀 Heart Disease (Classification)

**Goal:** Predict `AHD` (Yes/No heart disease) from 13+ clinical
predictors like age, cholesterol, resting blood pressure, etc.

- **Outcome:** binary (`Yes` / `No`)

```{r}
#| label: download-data
# heart <- read.csv("https://www.statlearning.com/s/Heart.csv",
#                   row.names = 1)

heart <- readr::read_csv("Day 2/Heart.csv") |>
  select(-1)

heart$AHD <- as.factor(heart$AHD)
heart <- na.omit(heart) ##remove NA
head(heart)
# # A tibble: 6 × 14
#     Age   Sex ChestPain    RestBP  Chol   Fbs RestECG MaxHR ExAng Oldpeak Slope    Ca Thal       AHD  
#   <dbl> <dbl> <chr>         <dbl> <dbl> <dbl>   <dbl> <dbl> <dbl>   <dbl> <dbl> <dbl> <chr>      <fct>
# 1    63     1 typical         145   233     1       2   150     0     2.3     3     0 fixed      No   
# 2    67     1 asymptomatic    160   286     0       2   108     1     1.5     2     3 normal     Yes  
# 3    67     1 asymptomatic    120   229     0       2   129     1     2.6     2     2 reversable Yes  
# 4    37     1 nonanginal      130   250     0       0   187     0     3.5     3     0 normal     No   
# 5    41     0 nontypical      130   204     0       2   172     0     1.4     1     0 normal     No   
# 6    56     1 nontypical      120   236     0       0   178     0     0.8     1     0 normal     No 
  
table(heart$AHD)
```


```{r}
#| label: summarize-data
heart |>
    summarize(
        across(everything(), ~mean(is.na(.x)), .names = "{.col}_p_miss")
    )
```

## Train-test Split

We will evaluate the performance of the fully trained/cross-validated models on this hold out test set (20\%) of the data.

I'll try setting up a recipe

```{r}
#| label: train-test-split
heart_split <- heart |> initial_split(prop = 0.8, strata = AHD)

heart_split

heart_rec <- recipe(
    AHD ~ .,
    data = training(heart_split)
) |>
    step_normalize(all_numeric_predictors()) |>
    prep()

heart_train <- bake(heart_rec, new_data = NULL)
heart_test <- bake(heart_rec, new_data = testing(heart_split))
```


```{r}
#| label: train-cv
train_cv <- vfold_cv(heart_train, v = 5)

train_cv
```

# Decision Tree w/ `rpart` via `parsnip`

```{r}
#| label: dt-spec
DT_spec <- decision_tree(
    cost_complexity = tune(),
    tree_depth = tune(),
    min_n = tune()
) |> # rpart is the default engine
    set_mode("classification")
```

```{r}
#| label: dt-grid
DT_grid <- grid_space_filling(
  tree_depth(),
  min_n(),
  cost_complexity(),
  size = 100
)

DT_grid
```

```{r}
#| label: dt-wf
DT_wf <- workflow() |>
  add_formula(AHD ~ .) |>
  add_model(DT_spec)

DT_wf
```

```{r}
#| label: dt-train-cv
doParallel::registerDoParallel(cores = 4)

set.seed(123958)

DT_res <- tune_grid(
  DT_wf,
  resamples = train_cv,
  grid = DT_grid,
  control = control_grid(save_pred = TRUE)
)

DT_res
```

```{r}
#| label: dt-cv-res
collect_metrics(DT_res)

DT_res |>
  collect_metrics() |>
  filter(.metric == "accuracy") |>
  select(mean, cost_complexity:min_n) |>
  pivot_longer(cost_complexity:min_n,
               values_to = "value",
               names_to = "parameter"
  ) |>
  ggplot(aes(value, mean, color = parameter)) +
  geom_point(alpha = 0.8, show.legend = FALSE) +
  facet_wrap(~parameter, scales = "free_x") +
  labs(x = NULL, y = "Accuracy")
```

```{r}
#| label: dt-best-met
show_best(DT_res, metric = "accuracy")
best_DT <- select_best(DT_res, metric = "accuracy")
best_DT
```

```{r}
#| label: dt-final
DT_final_wf <- finalize_workflow(
  DT_wf,
  best_DT
)
DT_final_wf

DT_final <- DT_final_wf |>
  fit(data = heart_train)

DT_final |>
  pull_workflow_fit() |> 
  vip(geom = "point")
```


```{r}
#| label: dt-pred
DT_final |> predict(type = "prob", new_data = heart_test)

DT_pred_train <- predict(DT_final, new_data = heart_train)
DT_pred_train_YN <- DT_pred_train$.pred_class |> factor()
DT_pred <- predict(DT_final, new_data = heart_test)
DT_pred_YN <- DT_pred$.pred_class |> factor()

# Confusion Matrix
conf_matrix_DT <- confusionMatrix(DT_pred_YN, heart_test$AHD)
print(conf_matrix_DT)
```

# Random Forest

```{r}
#| label: rf-spec
RF_spec <- rand_forest(
    mtry = tune(),
    trees = tune(),
    min_n = tune()
) |> 
    set_engine("ranger", keep.inbag = TRUE) |>
    set_mode("classification")
```

```{r}
#| label: rf-grid
RF_grid <- grid_space_filling(
  finalize(mtry(), heart_train),
  min_n(),
  trees(),
  size = 100
)

RF_grid
```

```{r}
#| label: rf-wf
RF_wf <- workflow() |>
  add_formula(AHD ~ .) |>
  add_model(RF_spec)

RF_wf
```

```{r}
#| label: rf-train-cv
doParallel::registerDoParallel(cores = 4)

set.seed(123958)

RF_res <- tune_grid(
  RF_wf,
  resamples = train_cv,
  grid = RF_grid,
  control = control_grid(save_pred = TRUE)
)

RF_res
```

```{r}
#| label: rf-cv-res
collect_metrics(RF_res)

RF_res |>
  collect_metrics() |>
  filter(.metric == "accuracy") |>
  select(mean, mtry:min_n) |>
  pivot_longer(mtry:min_n,
               values_to = "value",
               names_to = "parameter"
  ) |>
  ggplot(aes(value, mean, color = parameter)) +
  geom_point(alpha = 0.8, show.legend = FALSE) +
  facet_wrap(~parameter, scales = "free_x") +
  labs(x = NULL, y = "Accuracy")
```

```{r}
#| label: rf-best-met
show_best(RF_res, metric = "accuracy")
best_RF <- select_best(RF_res, metric = "accuracy")
best_RF
```

```{r}
#| label: rf-final
RF_final_wf <- finalize_workflow(
  RF_wf,
  best_RF
)
RF_final_wf

RF_final <- RF_final_wf |>
  fit(data = heart_train)
```


```{r}
#| label: rf-pred
RF_final |> predict(type = "prob", new_data = heart_test)

RF_pred_train <- predict(RF_final, new_data = heart_train)
RF_pred_train_YN <- RF_pred_train$.pred_class |> factor()
RF_pred <- predict(RF_final, new_data = heart_test)
RF_pred_YN <- RF_pred$.pred_class |> factor()

# Confusion Matrix
conf_matrix_RF <- confusionMatrix(RF_pred_YN, heart_test$AHD)
print(conf_matrix_RF)
```

# BART

```{r}
#| label: bart-spec
BART_spec <- bart(
    trees = tune(),
    prior_terminal_node_coef = tune(),
    prior_terminal_node_expo = tune(),
    prior_outcome_range = tune()
) |> # rpart is the default engine
    set_mode("classification")
```

```{r}
#| label: bart-grid
BART_grid <- grid_space_filling(
  trees(c(100, 1000)),
  prior_terminal_node_coef(range = c(0,1)),
  prior_terminal_node_expo(range = c(0,1)),
  prior_outcome_range(range = c(0,1)),
  size = 100
)

BART_grid
```

```{r}
#| label: bart-wf
BART_wf <- workflow() |>
  add_formula(AHD ~ .) |>
  add_model(BART_spec)

BART_wf
```

```{r}
#| label: bart-train-cv
doParallel::registerDoParallel(cores = 4)

set.seed(123958)

BART_res <- tune_grid(
  BART_wf,
  resamples = train_cv,
  grid = BART_grid,
  control = control_grid(save_pred = TRUE)
)

BART_res
```

```{r}
#| label: bart-res
collect_metrics(BART_res)

BART_res |>
  collect_metrics() |>
  filter(.metric == "accuracy") |>
  select(mean, trees:prior_outcome_range) |>
  pivot_longer(trees:prior_outcome_range,
               values_to = "value",
               names_to = "parameter"
  ) |>
  ggplot(aes(value, mean, color = parameter)) +
  geom_point(alpha = 0.8, show.legend = FALSE) +
  facet_wrap(~parameter, scales = "free_x") +
  labs(x = NULL, y = "Accuracy")
```

```{r}
#| label: bart-best-met
show_best(BART_res, metric = "accuracy")
best_BART <- select_best(BART_res, metric = "accuracy")
best_BART
```

```{r}
#| label: bart-final
BART_final_wf <- finalize_workflow(
  BART_wf,
  best_BART
)
BART_final_wf

BART_final <- BART_final_wf |>
  fit(data = heart_train)
```


```{r}
#| label: bart-pred
BART_final |> predict(type = "prob", new_data = heart_test)

BART_pred_train <- predict(BART_final, new_data = heart_train)
BART_pred_train_YN <- BART_pred_train$.pred_class |> factor()
BART_pred <- predict(BART_final, new_data = heart_test)
BART_pred_YN <- BART_pred$.pred_class |> factor()

# Confusion Matrix
conf_matrix_BART <- confusionMatrix(BART_pred_YN, heart_test$AHD)
print(conf_matrix_BART)
```



# Trying out XGBoost with `tidymodels`

## Tuning XGBoost with `tidymodels` and CV

```{r}
#| label: xgb-spec
xgb_spec <- boost_tree(
  trees = 1000,
  tree_depth = tune(),
  min_n = tune(),
  loss_reduction = tune(),
  sample_size = tune(),
  mtry = tune(),
  learn_rate = tune()
) |>
  set_engine("xgboost") |>
  set_mode("classification")

xgb_spec
```

```{r}
#| label: xgb-grid
xgb_grid <- grid_space_filling(
  tree_depth(),
  min_n(),
  loss_reduction(),
  sample_size = sample_prop(),
  finalize(mtry(), heart_train),
  learn_rate(),
  size = 100
)

xgb_grid
```

```{r}
#| label: xgb-wf
xgb_wf <- workflow() |>
  add_formula(AHD ~ .) |>
  add_model(xgb_spec)

xgb_wf
```

```{r}
#| label: xgb-train-cv
set.seed(123958)

xgb_res <- tune_grid(
  xgb_wf,
  resamples = train_cv,
  grid = xgb_grid,
  control = control_grid(save_pred = TRUE)
)

xgb_res
```

```{r}
#| label: xgb-met
collect_metrics(xgb_res)

xgb_res |>
  collect_metrics() |>
  filter(.metric == "accuracy") |>
  select(mean, mtry:sample_size) |>
  pivot_longer(mtry:sample_size,
               values_to = "value",
               names_to = "parameter"
  ) |>
  ggplot(aes(value, mean, color = parameter)) +
  geom_point(alpha = 0.8, show.legend = FALSE) +
  facet_wrap(~parameter, scales = "free_x") +
  labs(x = NULL, y = "Accuracy")
```

```{r}
#| label: xgb-best-met
show_best(xgb_res, metric = "accuracy")
best_auc <- select_best(xgb_res, metric = "accuracy")
best_auc

XGB_final_wf <- finalize_workflow(
  xgb_wf,
  best_auc
)
XGB_final_wf
```

```{r}
#| label: xgb-final
XGB_final <- XGB_final_wf |>
  fit(data = heart_train)

XGB_final |>
  pull_workflow_fit() |> 
  vip(geom = "point")
```


```{r}
#| label: xgb-pred
XGB_final |> predict(type = "prob", new_data = heart_test)


xgb_pred_train <- predict(XGB_final, new_data = heart_train)
xgb_pred_train_YN <- xgb_pred_train$.pred_class |> factor()
xgb_pred <- predict(XGB_final, new_data = heart_test)
xgb_pred_YN <- xgb_pred$.pred_class |> factor()

# Confusion Matrix
conf_matrix_xgb <- confusionMatrix(xgb_pred_YN, heart_test$AHD)
print(conf_matrix_xgb)
```

# Model Comparison

## Performance Summary Table

```{r}
#| label: tbl-comp
#| tbl-cap: "Test-Set Performance Comparison"
library(kableExtra)

# Extract metrics from confusion matrices
metrics_dt <- data.frame(
  Model = "Decision Tree",
  Accuracy = conf_matrix_DT$overall["Accuracy"],
  Sensitivity = conf_matrix_DT$byClass["Sensitivity"],
  Specificity = conf_matrix_DT$byClass["Specificity"],
  PPV = conf_matrix_DT$byClass["Pos Pred Value"],
  NPV = conf_matrix_DT$byClass["Neg Pred Value"]
)

metrics_rf <- data.frame(
  Model = "Random Forest",
  Accuracy = conf_matrix_RF$overall["Accuracy"],
  Sensitivity = conf_matrix_RF$byClass["Sensitivity"],
  Specificity = conf_matrix_RF$byClass["Specificity"],
  PPV = conf_matrix_RF$byClass["Pos Pred Value"],
  NPV = conf_matrix_RF$byClass["Neg Pred Value"]
)

metrics_bart <- data.frame(
  Model = "BART",
  Accuracy = conf_matrix_BART$overall["Accuracy"],
  Sensitivity = conf_matrix_BART$byClass["Sensitivity"],
  Specificity = conf_matrix_BART$byClass["Specificity"],
  PPV = conf_matrix_BART$byClass["Pos Pred Value"],
  NPV = conf_matrix_BART$byClass["Neg Pred Value"]
)

metrics_xgb <- data.frame(
  Model = "XGBoost",
  Accuracy = conf_matrix_xgb$overall["Accuracy"],
  Sensitivity = conf_matrix_xgb$byClass["Sensitivity"],
  Specificity = conf_matrix_xgb$byClass["Specificity"],
  PPV = conf_matrix_xgb$byClass["Pos Pred Value"],
  NPV = conf_matrix_xgb$byClass["Neg Pred Value"]
)

comparison_table <- rbind(metrics_dt, metrics_rf, metrics_bart, metrics_xgb)

comparison_table |>
  mutate(across(Accuracy:NPV, ~round(., 4))) |>
  kable(format = "html") |>
  kable_styling(bootstrap_options = c("striped", "hover"),
                full_width = FALSE) |>
  row_spec(which.max(comparison_table$Accuracy), bold = TRUE,
           color = "white", background = "#1a9641")
```

## Visual Comparison of Performance Metrics

```{r}
#| label: fig-comp
#| fig-cap: Model predictive performance comparison.
library(tidyr)
comparison_table_long <- comparison_table |>
  pivot_longer(cols = -Model, names_to = "Metric", values_to = "Value")

ggplot(comparison_table_long, aes(x = Metric, y = Value, fill = Model)) +
  geom_bar(stat = "identity", position = "dodge") +
  facet_wrap(~Metric, scales = "free") +
  theme_minimal(base_size = 13) +
  labs(title = "Model Performance Metrics Comparison", x = NULL, y = "Value") +
  theme(legend.position = "bottom")
```

---

# Exercises {.unnumbered}

## Student Exercise: Regularized Logistic Regression {.unnumbered}

**Goal:** Use a regularized logistic regression model (e.g., using `glmnet` via `tidymodels`) to classify the outcome `AHD`.

1. **Model Specification:** Define a `logistic_reg` model specification with `penalty` and `mixture` (for Lasso/Ridge) as tuning parameters.
2. **Workflow:** Create a workflow that includes the `heart_rec` recipe.
3. **Tuning:** Use `tune_grid()` to find the optimal `penalty` and `mixture` values using cross-validation.
4. **Final Model:** Train the final model on the training set and evaluate its performance on the test set.


```{r}
# 1. Model Specification
# hint: Use `logistic_reg()` and `set_engine("glmnet")`. 
# Use `tune()` for `penalty` and `mixture`.
```

```{r}
# 2. Workflow
# hint: Use `workflow()` and `add_recipe(heart_rec)`.
```

```{r}
# 3. Tuning
# hint: Use `vfold_cv()` to create resamples, then `tune_grid()`.
```

```{r}
# 4. Final Model
# hint: Use `finalize_workflow()` and `last_fit()`.
```
