---
title: "Tidymodels for Classification"
author: "Ben Stockton"
format: 
    html:
        theme: flatly
        toc: true
        toc-location: left
        code-fold: false
        highlight-style: github
        fig-align: center
        number-sections: true
    revealjs:
        output-file: tree_class_slides
        theme: simple
        slide-number: true
        chalkboard: true
        code-fold: false
        highlight-style: github
        fig-align: center
        incremental: false
# bibliography: references.bib
# csl: https://raw.githubusercontent.com/citation-style-language/styles/master/apa.csl
execute:
  echo: true
  warning: false
  message: false
  cache: true
---

# Introduction

This lab explores tree based methods - decision trees and random forest - for classification, i.e. prediction of binary outcome.

## Set-up

```{r}
library(dplyr)
library(ggplot2)
library(caret)
library(randomForest)
library(rpart)
library(rpart.plot)
library(tidymodels)
library(vip)
```


---

## 🫀 Heart Disease (Classification)

**Goal:** Predict `AHD` (Yes/No heart disease) from 13+ clinical
predictors like age, cholesterol, resting blood pressure, etc.

- **Outcome:** binary (`Yes` / `No`)

```{r}
#| label: download-data
# heart <- read.csv("https://www.statlearning.com/s/Heart.csv",
#                   row.names = 1)

heart <- readr::read_csv("Day 2/Heart.csv") |>
  select(-1)

heart$AHD <- as.factor(heart$AHD)
heart <- na.omit(heart) ##remove NA
head(heart)
# # A tibble: 6 × 14
#     Age   Sex ChestPain    RestBP  Chol   Fbs RestECG MaxHR ExAng Oldpeak Slope    Ca Thal       AHD  
#   <dbl> <dbl> <chr>         <dbl> <dbl> <dbl>   <dbl> <dbl> <dbl>   <dbl> <dbl> <dbl> <chr>      <fct>
# 1    63     1 typical         145   233     1       2   150     0     2.3     3     0 fixed      No   
# 2    67     1 asymptomatic    160   286     0       2   108     1     1.5     2     3 normal     Yes  
# 3    67     1 asymptomatic    120   229     0       2   129     1     2.6     2     2 reversable Yes  
# 4    37     1 nonanginal      130   250     0       0   187     0     3.5     3     0 normal     No   
# 5    41     0 nontypical      130   204     0       2   172     0     1.4     1     0 normal     No   
# 6    56     1 nontypical      120   236     0       0   178     0     0.8     1     0 normal     No 
  
table(heart$AHD)
```


```{r}
heart |>
    summarize(
        across(everything(), ~mean(is.na(.x)), .names = "{.col}_p_miss")
    )
```

## Train-test Split

We will evaluate the performance of the fully trained/cross-validated models on this hold out test set (20\%) of the data.

I'll try setting up a recipe

```{r}
heart_split <- heart |> initial_split(prop = 0.8, strata = AHD)

heart_split

heart_rec <- recipe(
    AHD ~ .,
    data = training(heart_split)
) |>
    step_normalize(all_numeric_predictors()) |>

heart_train <- bake(heart_rec, new_data = NULL)
heart_test <- bake(heart_rec, new_data = testing(heart_split))
```


```{r}
train_cv <- vfold_cv(heart_train, v = 5)

train_cv
```

# Decision Tree w/ `rpart` via `parsnip`

```{r}
DT_spec <- decision_tree(
    cost_complexity = tune(),
    tree_depth = tune(),
    min_n = tune()
) |> # rpart is the default engine
    set_mode("classification")
```

```{r}
DT_grid <- grid_space_filling(
  tree_depth(),
  min_n(),
  cost_complexity(),
  size = 100
)

DT_grid
```

```{r}
DT_wf <- workflow() |>
  add_formula(AHD ~ .) |>
  add_model(DT_spec)

DT_wf
```

```{r}
doParallel::registerDoParallel(cores = 4)

set.seed(123958)

DT_res <- tune_grid(
  DT_wf,
  resamples = train_cv,
  grid = DT_grid,
  control = control_grid(save_pred = TRUE)
)

DT_res
```

```{r}
collect_metrics(DT_res)

DT_res |>
  collect_metrics() |>
  filter(.metric == "accuracy") |>
  select(mean, cost_complexity:min_n) |>
  pivot_longer(cost_complexity:min_n,
               values_to = "value",
               names_to = "parameter"
  ) |>
  ggplot(aes(value, mean, color = parameter)) +
  geom_point(alpha = 0.8, show.legend = FALSE) +
  facet_wrap(~parameter, scales = "free_x") +
  labs(x = NULL, y = "Accuracy")
```

```{r}
show_best(DT_res, metric = "accuracy")
best_DT <- select_best(DT_res, metric = "accuracy")
best_DT
```

```{r}
DT_final_wf <- finalize_workflow(
  DT_wf,
  best_DT
)
DT_final_wf

DT_final <- DT_final_wf |>
  fit(data = heart_train)

DT_final |>
  pull_workflow_fit() |> 
  vip(geom = "point")
```


```{r}
DT_final |> predict(type = "prob", new_data = heart_test)
```

```{r}
DT_pred_train <- predict(DT_final, new_data = heart_train)
DT_pred_train_YN <- DT_pred_train$.pred_class |> factor()
DT_pred <- predict(DT_final, new_data = heart_test)
DT_pred_YN <- DT_pred$.pred_class |> factor()

# Confusion Matrix
conf_matrix_DT <- confusionMatrix(DT_pred_YN, heart_test$AHD)
print(conf_matrix_DT)
```

# Random Forest

```{r}
RF_spec <- rand_forest(
    mtry = tune(),
    trees = tune(),
    min_n = tune()
) |> 
    set_engine("ranger", keep.inbag = TRUE) |>
    set_mode("classification")
```

```{r}
RF_grid <- grid_space_filling(
  finalize(mtry(), heart_train),
  min_n(),
  trees(),
  size = 100
)

RF_grid
```

```{r}
RF_wf <- workflow() |>
  add_formula(AHD ~ .) |>
  add_model(RF_spec)

RF_wf
```

```{r}
doParallel::registerDoParallel(cores = 4)

set.seed(123958)

RF_res <- tune_grid(
  RF_wf,
  resamples = train_cv,
  grid = RF_grid,
  control = control_grid(save_pred = TRUE)
)

RF_res
```

```{r}
collect_metrics(RF_res)

RF_res |>
  collect_metrics() |>
  filter(.metric == "accuracy") |>
  select(mean, mtry:min_n) |>
  pivot_longer(mtry:min_n,
               values_to = "value",
               names_to = "parameter"
  ) |>
  ggplot(aes(value, mean, color = parameter)) +
  geom_point(alpha = 0.8, show.legend = FALSE) +
  facet_wrap(~parameter, scales = "free_x") +
  labs(x = NULL, y = "Accuracy")
```

```{r}
show_best(RF_res, metric = "accuracy")
best_RF <- select_best(RF_res, metric = "accuracy")
best_RF
```

```{r}
RF_final_wf <- finalize_workflow(
  RF_wf,
  best_RF
)
RF_final_wf

RF_final <- RF_final_wf |>
  fit(data = heart_train, 
      )
```


```{r}
RF_final |> predict(type = "prob", new_data = heart_test)
```

```{r}
RF_pred_train <- predict(RF_final, new_data = heart_train)
RF_pred_train_YN <- RF_pred_train$.pred_class |> factor()
RF_pred <- predict(RF_final, new_data = heart_test)
RF_pred_YN <- RF_pred$.pred_class |> factor()

# Confusion Matrix
conf_matrix_RF <- confusionMatrix(RF_pred_YN, heart_test$AHD)
print(conf_matrix_RF)
```

# BART

```{r}
BART_spec <- bart(
    trees = tune(),
    prior_terminal_node_coef = tune(),
    prior_terminal_node_expo = tune(),
    prior_outcome_range = tune()
) |> # rpart is the default engine
    set_mode("classification")
```

```{r}
BART_grid <- grid_space_filling(
  trees(c(100, 1000)),
  prior_terminal_node_coef(range = c(0,1)),
  prior_terminal_node_expo(range = c(0,1)),
  prior_outcome_range(range = c(0,1)),
  size = 100
)

BART_grid
```

```{r}
BART_wf <- workflow() |>
  add_formula(AHD ~ .) |>
  add_model(BART_spec)

BART_wf
```

```{r}
doParallel::registerDoParallel(cores = 4)

set.seed(123958)

BART_res <- tune_grid(
  BART_wf,
  resamples = train_cv,
  grid = BART_grid,
  control = control_grid(save_pred = TRUE)
)

BART_res
```

```{r}
collect_metrics(BART_res)

BART_res |>
  collect_metrics() |>
  filter(.metric == "accuracy") |>
  select(mean, trees:prior_outcome_range) |>
  pivot_longer(trees:prior_outcome_range,
               values_to = "value",
               names_to = "parameter"
  ) |>
  ggplot(aes(value, mean, color = parameter)) +
  geom_point(alpha = 0.8, show.legend = FALSE) +
  facet_wrap(~parameter, scales = "free_x") +
  labs(x = NULL, y = "Accuracy")
```

```{r}
show_best(BART_res, metric = "accuracy")
best_BART <- select_best(BART_res, metric = "accuracy")
best_BART
```

```{r}
BART_final_wf <- finalize_workflow(
  BART_wf,
  best_BART
)
BART_final_wf

BART_final <- BART_final_wf |>
  fit(data = heart_train)

BART_final |>
  pull_workflow_fit() |> 
  vip(geom = "point")
```


```{r}
BART_final |> predict(type = "prob", new_data = heart_test)
```

```{r}
BART_pred_train <- predict(BART_final, new_data = heart_train)
BART_pred_train_YN <- BART_pred_train$.pred_class |> factor()
BART_pred <- predict(BART_final, new_data = heart_test)
BART_pred_YN <- BART_pred$.pred_class |> factor()

# Confusion Matrix
conf_matrix_BART <- confusionMatrix(BART_pred_YN, heart_test$AHD)
print(conf_matrix_BART)
```



# Trying out XGBoost with `tidymodels`

## Tuning XGBoost with `tidymodels` and CV

```{r}
xgb_spec <- boost_tree(
  trees = 1000,
  tree_depth = tune(),
  min_n = tune(),
  loss_reduction = tune(),
  sample_size = tune(),
  mtry = tune(),
  learn_rate = tune()
) |>
  set_engine("xgboost") |>
  set_mode("classification")

xgb_spec
```

```{r}
xgb_grid <- grid_space_filling(
  tree_depth(),
  min_n(),
  loss_reduction(),
  sample_size = sample_prop(),
  finalize(mtry(), heart_train),
  learn_rate(),
  size = 100
)

xgb_grid
```

```{r}
xgb_wf <- workflow() |>
  add_formula(AHD ~ .) |>
  add_model(xgb_spec)

xgb_wf
```

```{r}
set.seed(123958)

xgb_res <- tune_grid(
  xgb_wf,
  resamples = train_cv,
  grid = xgb_grid,
  control = control_grid(save_pred = TRUE)
)

xgb_res
```

```{r}
collect_metrics(xgb_res)
```

```{r}
xgb_res |>
  collect_metrics() |>
  filter(.metric == "accuracy") |>
  select(mean, mtry:sample_size) |>
  pivot_longer(mtry:sample_size,
               values_to = "value",
               names_to = "parameter"
  ) |>
  ggplot(aes(value, mean, color = parameter)) +
  geom_point(alpha = 0.8, show.legend = FALSE) +
  facet_wrap(~parameter, scales = "free_x") +
  labs(x = NULL, y = "Accuracy")
```

```{r}
show_best(xgb_res, metric = "accuracy")
best_auc <- select_best(xgb_res, metric = "accuracy")
best_auc
```

```{r}
XGB_final_wf <- finalize_workflow(
  xgb_wf,
  best_auc
)
XGB_final_wf
```

```{r}
XGB_final <- XGB_final_wf |>
  fit(data = heart_train)

XGB_final |>
  pull_workflow_fit() |> 
  vip(geom = "point")
```


```{r}
XGB_final |> predict(type = "prob", new_data = heart_test)
```

```{r}
xgb_pred_train <- predict(XGB_final, new_data = heart_train)
xgb_pred_train_YN <- xgb_pred_train$.pred_class |> factor()
xgb_pred <- predict(XGB_final, new_data = heart_test)
xgb_pred_YN <- xgb_pred$.pred_class |> factor()

# Confusion Matrix
conf_matrix_xgb <- confusionMatrix(xgb_pred_YN, heart_test$AHD)
print(conf_matrix_xgb)
```


# Model Comparison

## Visualise Predictions vs. Actuals

```{r}
# Proportion of cases
pred_dt_train <- predict(tree_pruned)
head(pred_dt_train) ##note the output is predicted probabilityes
pred_dt_train <- ifelse(pred_dt_train[,2] > 0.5, "Yes", "No")
pred_rf_train <- predict(rf_model)

pred_DT <- tibble(
    Actual = c(heart_train$AHD, heart_test$AHD),
    Pred = c(pred_dt_train, tree_preds_YN),
    split = c(rep("train", nrow(heart_train)), rep("test", nrow(heart_test))),
    model = "DT"
)

pred_RF <- tibble(
    Actual = c(heart_train$AHD, heart_test$AHD),
    Pred = c(pred_rf_train, rf_preds),
    split = c(rep("train", nrow(heart_train)), rep("test", nrow(heart_test))),
    model = "RF"
)

pred_XGB <- tibble(
    Actual = c(heart_train$AHD, heart_test$AHD),
    Pred = c(xgb_pred_train_YN, xgb_pred_YN),
    split = c(rep("train", nrow(heart_train)), rep("test", nrow(heart_test))),
    model = "XGBoost"
)
pred_df <- bind_rows(pred_DT, pred_RF, pred_XGB) |>
    print(n = 5)
```

---

```{r}
pred_metrics <- pred_df |>
    group_by(split, model) |>
    summarize(
        accuracy = mean(Actual == Pred),
        sensitivity = sum(Actual == "Yes" & Pred == "Yes") / sum(Actual == "Yes"),
        specificity = sum(Actual == "No" & Pred == "No") / sum(Actual == "No"),
        pos_pred_value = sum(Actual == "Yes" & Pred == "Yes") / sum(Pred == "Yes"),
        neg_pred_value = sum(Actual == "No" & Pred == "No") / sum(Pred == "No")
    ) |>
    print() |>
    tidyr::pivot_longer(
        accuracy:neg_pred_value,
        names_to = "metric",
        values_to = "value"
    )
```

---

```{r}
# Plot proportions
ggplot(pred_df, aes(x = Actual, fill = Pred)) +
  geom_bar(position = "dodge") +
  scale_fill_discrete(name = "Pred") +
  labs(
    title    = "Random Forest: Proportions of Predicted vs Actual AHD",
    x = "Actual HD", y = "Proportion"
  ) +
  theme_classic() +
  facet_wrap(split + model~.) +
  theme_minimal(base_size = 14)
```

---

```{r}
ggplot(pred_metrics, aes(metric, value, color = model)) +
    geom_point(aes(shape = split), size = 2) +
    facet_wrap(.~split, nrow = 2) +
  theme_minimal(base_size = 14) +
  theme(axis.text.x = element_text(angle = 45, hjust = 1))

```


## Performance Summary Table

```{r comparison-table}
library(kableExtra)

pred_met_wide <- pred_df |>
  filter(split == "test") |>
    group_by(model) |>
    summarize(
        accuracy = mean(Actual == Pred),
        sensitivity = sum(Actual == "Yes" & Pred == "Yes") / sum(Actual == "Yes"),
        specificity = sum(Actual == "No" & Pred == "No") / sum(Actual == "No"),
        pos_pred_value = sum(Actual == "Yes" & Pred == "Yes") / sum(Pred == "Yes"),
        neg_pred_value = sum(Actual == "No" & Pred == "No") / sum(Pred == "No")
    ) 

pred_met_wide |>
  mutate(across(c(accuracy, sensitivity), ~ round(., 4))) |>
  kable(caption = "Test-Set Performance — All Models") |>
  kable_styling(bootstrap_options = c("striped", "hover"),
                full_width = FALSE) |>
  row_spec(which.max(pred_met_wide$accuracy), bold = TRUE,
           color = "white", background = "#1a9641")
```

## RMSE Comparison Plot

```{r rmse-plot, fig.height=4}
pred_met_wide |>
  mutate(Model = factor(model, levels = model[order(accuracy)])) |>
  ggplot(aes(x = model, y = accuracy, fill = model)) +
  geom_hline(yintercept = 1, color = "gray50", linetype = "dashed") +
  geom_col(alpha = 0.85, width = 0.6) +
  geom_text(aes(label = round(accuracy, 3)), hjust = -0.1, size = 3.8) +
  coord_flip() +
  scale_fill_brewer(palette = "RdYlBu") +
  labs(title = "Test-Set accuracy by Model",
       x = NULL, y = "accuracy") +
  theme_minimal(base_size = 13) +
  theme(legend.position = "none") +
  ylim(0, 1.05)
```
